Can historical data predict population responses to climate change experiments?

### 2016-12-16 precip/manuscript/precip_ms.tex
adding edits to the introduction of the manuscript

\begin{enumerate}
\item Climate is thought to be an important driver of population ecology however it can be difficult to determine which aspects of climate affect populations and at which stages in an organisms life-cycle.  Moreover, there have been few tests of whether statistical models fit to observational data on climate and populations can accurately predict the future responses of populations to natural or experimentally imposed climate conditions.   
\item Using observational data collected from the 1920's to the 1950's from a sagebrush steppe plant community, previous research has reported a wide variety of climate effects on four dominant plant species. We tested whether statistical models based on these historical data could predict how these species respond to natural and experimentally imposed climate variation from 2006 to 2016, some 60 years after the historical data collection ended. 
\\item We established sixteen new plots in the same location as the historical data and assigned each to either drought or irrigation treatments imposed using rainfall shelters and automatic sprinklers. The original plots were also monitored and served as ambient climate controls. Using the historical data, we fit statistical models for survival, growth and recruitment of each species in response to seasonal temperature and soil moisture.  We used these models to predict variation in species vital rates in the experiment and used a simple individual based population model to predict one year ahead population growth rates in each of the experimental plots.   
\item Over the course of the experiment, average plot cover of \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined, while cover of \textit{Poa secunda} showed a trend towards increase in the irrigated plots. At the level of individual vital rates, experimental drought led to decreases in \textit{} growth, survival, while irrigation led to increases in \textit{}. 
\item Vital rate models fit to the historical data correctly predicted significant increases in X and X and significant decreaes in X and X.  However, the historical models predicted significant changes in X X and X that were not observed in the experiment.  Moreover, the experiment led to a significant change in X that was not predicted by the historical data. 
\item Our individual based population model more accurately predicted changes in the population growth rates of \textit{} in the experimental plots than a model without climate effects.  However for the other X species predictions made from the climate model were less accurate than models without climate.   
\item \emph{Synthesis}: Out of all four species examined, climate covariates only appeared to increase our predictive accuracy for one species: \textit{Hesperostipa comata}. In general the relatively weak responses to the experimental treatments agrees with the low explanatory power of climate covariates in explaining the vital rates and population dynamics of these species in this system. The highly variable environment and water limited conditions may select for species with a conservative non-responsive strategy to variation in climate both natural and experimentally imposed.

@@

Climate is widely considered one of the most powerful external forces driving changes in species abundance across space and time \citep{teller_linking_2016}.  The effects of climate on populations and ecosystems are most apparent at the largest scales in space and time: climate determines the distribution of ecosystems, treelines and the the range limits of many species \citep{}.  The paleoecological record shows that long-term climate change lead to changes in species abundance and changes in species' ranges. Nevertheless, for many species it is difficult to determine how annual climate variation affects their populations and demographic rates \citep{ehrlen_advancing_2016}. Observational data on species performance across many different years can provide some information on how climate might affect population abundance \citep{dalgleish_climate_2010,ehrlen_advancing_2016}. But many years of data are needed to reliably detect climate effects, especially when climate effects are weak and often correlated amongst themselves \citep{teller_linking_2016,gerber_optimal_2015}. Understanding the effects of climate on populations is increasingly important, however, if we are to understand and predict the effects of climate change on populations. If models fit to historical data can be used to accurately predict the effects of contemporary climate variation on populations, as well as the effects of experimental climate manipulations, it would be strong confirmation that the climate effects they describe are real.

>>>>

\begin{enumerate}
\item Climate is thought to be an important driver of population ecology however it can be difficult to determine which aspects of climate affect populations and at which stages in an organisms life-cycle.  Moreover, there have been few tests of whether statistical models fit to observational data on climate and populations can accurately predict the future responses of populations to natural or experimentally imposed climate conditions.   
\item Using observational data collected from the 1920's to the 1950's from a sagebrush steppe plant community, previous research has reported a wide variety of climate effects on four dominant plant species. We tested whether statistical models based on these observational data could predict how these species respond to a experimentally imposed drought and irrigation from 2006 to 2016. 
\\item We established sixteen new plots in the same location as the historical data and assigned each to either drought or irrigation treatments imposed using rainfall shelters and automatic sprinklers. The original plots were also monitored and served as ambient climate controls. In order to describe the demographic effects of the climate manipulations, we used all data, observational and experimental, to fit a set of statistical models describing each species survival, growth and recruitment in response to local competitive interactions and the effects of the climate manipulation experiment. Next we fit another set of models to only the observational data that used temperature and soil moisture as covariates to describe annual variation in the vital rates.  Finally, we used the second set of models fitted only to the observational data, to predict the effects of the experimental treatments on each species vital rates and used the vital rates to parameterize an individual based population model and predict one step ahead cover in each of the experimental plots.   
\item The experimental drought and irrigation treatments successfully lowered and increased soil moisture respectively.  Over the course of the experiment, average plot cover of \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined, while cover of \textit{Poa secunda} showed a trend towards increase in the irrigated plots. At the level of individual vital rates, experimental drought reduced \textit{Hesperostipa comata} and \textit{Poa secunda} growth and survival. Drought increased and irrigation decreased \textit{Artemisia tripartita} growth. Recruitment rates for all three grasses were lower in both the drought and irrigation plots than in the control plots. 
\item Vital rate models fit to the observational data correctly predicted significant increases in X and X and significant decreaes in X and X.  However, the historical models predicted significant changes in X X and X that were not observed in the experiment.  Moreover, the experiment led to a significant change in X that was not predicted by the historical data. 
\item Our individual based population model more accurately predicted changes in the population growth rates of \textit{} in the experimental plots than a model without climate effects.  However for the other X species predictions made from the climate model were less accurate than models without climate.   
\item \emph{Synthesis}: Out of all four species examined, climate covariates only appeared to increase our predictive accuracy for one species: \textit{Hesperostipa comata}. In general the relatively weak responses to the experimental treatments agrees with the low explanatory power of climate covariates in explaining the vital rates and population dynamics of these species in this system. The highly variable environment and water limited conditions may select for species with a conservative non-responsive strategy to variation in climate both natural and experimentally imposed.

@@

Climate is widely considered one of the most powerful external forces driving changes in species abundance across space and time \citep{teller_linking_2016}.  The effects of climate on populations and ecosystems are most apparent at the largest scales in space and time: climate determines the distribution of ecosystems, treelines and the the range limits of many species \citep{}, while The paleoecological record shows that long-term climate change lead to changes in species abundance and changes in species' ranges \citep{davis}. Nevertheless, for many species it is difficult to determine how annual climate variation affects populations and demographic rates \citep{ehrlen_advancing_2016}. Observational data on species performance across years with varying climate can provide some information on how climate might affect population abundance \citep{dalgleish_climate_2010,ehrlen_advancing_2016}. However, many years of data are needed to reliably detect climate effects, especially when annual variation in demographic rates is high \citep{teller_linking_2016,gerber_optimal_2015}. Understanding the effects of climate on populations is increasingly important, however, if we are to understand and predict the effects of climate change on populations. If models fit to historical data can be used to accurately predict the effects of experimental climate manipulations, it would be strong confirmation that the climate effects they describe are real and will hold in the future \citep{}.

Many plant species occur across a wide range of climates \cite{} and individual plants in many environments must tolerate large fluctuations in seasonal temperature and soil moisture.   Unlike most animals, plants must endure these conditions in place and often become dormant in less optimal seasons \textit{}. At first pass, these observation would seem to indicate that annual variation in climate should have little effect on plant population performance.

Nevertheless, there is abundant evidence that plant performance shows high year to year variation, both at the level of individual growth, survival and reproduction and in terms of total population abundancend often this variation c. In many regions interannual variation in precipitation can be directly linked to variation in net primary productivity \citep{knapp and smith, Hsu and Adler}. The growth rates of trees are also often tightly linked to annual precipitation, so much so that annual growth rings in their stems can serve as accurate records of historical climate thousands of years before the present \citep{yang}.  Likewise, many smaller shrub and sub-shrub species show strong variation in growth in response to climate that is recorded in their tissues \citep{Franklin, Srur}. Among annual plants in desert ecosystems, germination and reproductive output is tightly linked to interannual precipitation variation \citep{venables}.

Linking annual climate variation to demographic performance in plants is a high priority for building population models for plants that can predict their future response to climate change \citep{crone_ability_2013,ehrlen_advancing_2016}. However, despite the clear signs that climate drives net primary productivity at the ecosystem level as and variation in individual growth rate in trees and other woody species, as well as the germination and fecundity of annual plants, there are few studies that have linked the effects of interannual climate variation to population models for plants, and fewer still that have shown that these population models can be used to accurately predict future plant responses to short term climate variation. Adler et al. \citep{adler byrn and leiker} showed that population models based on historically observed correlations between plant population growth rates and precipitation did have some predictive power in describing species response to a short-term climate manipulation in a North American grassland. Three species showed responses to experimentally imposed drought and irrigation that were well predicted by population models fitted to historical observations.  However, another three species, showed responses to the experimental conditions that were not well predicted by historical observations.  The authors suggested that limited replication in the historical data for two of these species and changing competitive conditions in the community may have led to the poor predictions.

The effects of annual climate variation on species in a sagebrush steppe plant community at the US Sheep Experiment Station near Dubois, Idaho have been described in at least X different studies since

Section on climate models in this system.
